---
title: "Whole Game Application"
author: "Soban Hassan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction
This report analyzes shipment data to answer key business questions using R. The analysis includes cleaning, transforming, and visualizing the data to identify trends and insights.

Project Plan

Plan Overview

The goal of this project is to analyze shipment data to understand how delivery time, shipment size, and other factors affect shipping costs and efficiency. The analysis will answer specific business questions about shipment costs, carrier performance, and the relationship between delivery time and cost.

Steps to Complete the Project
Import and Clean Data
The raw data will be imported into RStudio and cleaned by removing unnecessary columns, handling missing data, and renaming variables for consistency.

Data Transformation and Analysis
The cleaned data will be processed using dplyr functions such as mutate, filter, arrange, and group_by. Statistical summaries and visualizations will be created to identify patterns and relationships.

Reporting and Interpretation
The results will be presented in the form of tables and visualizations. The key findings will address the business questions and provide actionable insights into shipment costs and efficiency.

```{r}
# Load required libraries
library(tidyverse)
library(dplyr)
library(readxl)

# Import data
file_path <- "C:/git/Data332/Whole Game Raw Data.xlsx"
data <- read_excel(file_path)

# Preview data
head(data)
```

```{r}
# Remove unnecessary columns
data <- data %>%
  select(-`...10`)

# Rename columns for consistency
data <- data %>%
  rename(
    shipment_id = number,
    length = L,
    width = `W...3`,
    height = H,
    weight = `W...5`,
    carrier = carier,
    distance = `dist.`,
    cost = `($)`,
    delivery_days = days,
    diagonal = `longest diagonal`,
    density = `KG/m^3`,
    weight_distance = `KG-KM`
  )

# Handle missing values
data <- data %>%
  mutate(cost = replace_na(cost, 0))

# Display cleaned data
head(data)
```

```{r}
# Are shorter Delivery Times actually more expensive?
shorter_time_cost <- data %>%
  ggplot(aes(x = delivery_days, y = cost)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship between Delivery Time and Cost")

shorter_time_cost
```

```{r}
# How many shipments are completed by each Carrier?
shipments_per_carrier <- data %>%
  count(carrier)

shipments_per_carrier
```

```{r}
# What is average cost per shipment from each Carrier?
avg_cost_per_carrier <- data %>%
  group_by(carrier) %>%
  summarise(avg_cost = mean(cost, na.rm = TRUE))

avg_cost_per_carrier
```

```{r}
# What is cost per KG-KM for all Carriers?
cost_per_kg_km <- data %>%
  mutate(cost_per_kg_km = cost / weight_distance) %>%
  select(carrier, cost_per_kg_km)

cost_per_kg_km
```

```{r}
# What is the rank for each Carrier by Cost per KG-KM?
ranked_carriers <- cost_per_kg_km %>%
  group_by(carrier) %>%
  summarise(avg_cost_per_kg_km = mean(cost_per_kg_km, na.rm = TRUE)) %>%
  arrange(avg_cost_per_kg_km)

ranked_carriers
```

```{r}
# How does Longest Diagonal affect Cost?
diagonal_cost <- data %>%
  ggplot(aes(x = diagonal, y = cost)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship between Longest Diagonal and Cost")

diagonal_cost
```

```{r}
# 
final_data <- data %>%
  select(
    shipment_id, length, width, height, weight, carrier, 
    distance, cost, delivery_days, diagonal, volume, 
    density, weight_distance
  )

final_data
```

Conclusion

This analysis provides insights into shipment costs, carrier performance, and the relationship between delivery times and costs. The visualizations and summary tables clearly show the trends and key takeaways.



